Measuring Hash Trustworthiness via Collision Utility Metrics: Logical Cryptanalysis of MD4

### 2017-12-03 dsn-2018/bare_conf.tex
Changes suggested by KMayo
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

The remainder of this paper is organized as follows. In Section
\ref{Sec:Related} we discuss previous results and show how our research
expands upon prior work. In Section \ref{Sec:Terminology}, we discuss the
terminology and notations we use throughout the remainder of the paper.
In Section \ref{Sec:Intuition}, we give the intuition between measuring
the utility of a collision and motivate why it is useful. In Section
\ref{Sec:Logical}, we present new techniques in logical cryptanalysis to
analyze collisions in hash functions. In Section \ref{Sec:Empirical}, we
discuss the utility of existing collisions under the metrics described in
\ref{Sec:Intuition}. In Section \ref{Sec:Collision}, we discuss new
differential paths and the effectiveness of various techniques presented
in Section \ref{Sec:Logical}. Lastly, in Section \ref{Sec:Impacts}, we
discuss broader impacts of this research.

@@

Differential Cryptanalysis has been the major technique behind the discovery
of collisions in cryptographic hash functions. Its importance can be seen
everywhere from X. Wang's attacks on MD4 \cite{Wang2005} to more recent
cryptanalysis of general purpose hash functions such as Murmur3 by
J. Aumasson, D. Bernstein, and M. Bo{\ss}let \cite{murmur3DC}. However, while
differential cryptanalysis plays an important role, techniques for automated
analysis of hash functions are still an emerging area. Further, techniques
are being developed to make hash functions impervious to differential
analysis.

@@

Logical Cryptanalysis and its encoding as SAT likely started under the work of
F. Massacci in 1999 with his paper titled ``Using walk-SAT and rel-sat for
cryptographic key search''. \cite{Massacci:1999:UWR:1624218.1624261}. However,
much of the early work by F. Massacci was focused on symmetric and asymmetric
ciphers
\cite{Massacci:1999:UWR:1624218.1624261, MASSACCI2000LogicalCA, FIORINI2003101}.
It wasn't until the work of D. Jovanovi{\'{c}} and P. Jani{\v{c}}i{\'{c}}
that exploring collisions in the context of SAT was introduced
\cite{LogicalAnalysis}, and until the work of I. Mironov and L. Zhang that
this was studied for an existing collision class \cite{Mironov2006}.

@@

We make use of the \textit{bc2cnf} utility by T. Junttila as the basis for
our models \cite{circuits}. This allows us to encode hash functions as generic
structures in the circuit description language and ease the algorithmic
creation of models which build on top of them. We then use CryptoMiniSat5 by
M. Soos to run the models and check for a satisfying witness, often an
example collision \cite{Soos:2009:ESS:1575471.1575502, CryptoMiniSat5}. We
deviate from evaluating the performance of SAT solvers or encodings of models
by instead focusing on evaluating the properties of the hash function. We have
developed new techniques for using a SAT solver to inductively build full
collisions in MD4 from collisions in reduced-round MD4 and to build addition
examples of collision classes from a single instance of a collision in MD4.

@@

We can define a similar distance function, $\Delta$, between families of
collision classes by cardinality of the symmetric difference in the two
collision families. That is, given two families of collision classes,
$F_1, F_2 \in \mathcal{F}$:
\begin{align}
    \Delta(F_1, F_2) = \big| (F_1 \cup F_2) \setminus (F_1 \cap F_2) \big|
\end{align}
This is convenient for when the specifics of the differential path do not
matter, merely that there exist at least one collision class of the specified
form.

@@

Neighborhoods can be classified into three types: \textit{expansion},
\textit{internal}, and \textit{mixed}. Let $d$ be fixed. An \textit{expansion}
neighborhood of a collision class, $C \in \mathcal{C}$, is the neighborhood
restricted only to those collision classes which only differ in rounds external
to the collision family of $C$. An \textit{internal} neighborhood of $C$ is the
neighborhood restricted only to those collision classes which differ in rounds
internal to the collision family of $C$. An \textit{mixed} neighborhood of $C$
is the neighborhood restricted only to collisions which differ in a round
internal and a round external to the collision family of $C$. That is:
\begin{align*}
    N_{exp}(C, d) = \{ & C_{j} \in N(C, d) : \forall i \in F(C), C_j[i] = C[i] \} \\
    N_{int}(C, d) = \{ & C_{j} \in N(C, d) : \forall i \not\in F(C), C_j[i] = C[i] \} \\
    N_{mix}(C, d) = \{ & C_{j} \in N(C, d) : \exists i \in F(C), C_j[i] \neq C[i] \\
        & \text{ and } \exists k \not\in F(C), C_j[k] \neq C[k] \}
\end{align*}

@@

To begin, real world systems are most vulnerable to two attacks: the preimage
and second preimage. For systems relying on e.g., password validation via
comparing two hashes, preimage attacks would allow inverting the hash,
increasing the risk associated with losing a database of hashed passwords.
On the other hand, systems relying on message validation, such as signature
checks or file integrity checks would be easily broken with fast second
preimage techniques. Both of these attacks are intractable, thus protecting
real-world systems. However, much research has been devoted to collision
attacks, producing efficient results.

@@

Seeing the need to evaluate how much flexibility exists in a collision,
we propose the following metrics for evaluating the utility of a collision
class:
\begin{enumerate}
    \item The number of unique differentials a collision class has.
    \item The number of unit-step neighbors a collision class has.
    \item The maximum count of zeros in a the binary representation of a
        colliding block (and likewise with ones).
    \item Whether there exists a block which collides under multiple initial
        values.
    \item Whether or not zero, one, or both of the blocks in a collision may
        be of ASCII values under any input block difference.
\end{enumerate}
Note that the first three are quantitative measures providing some measure of
flexibility of a collision class, whereas the latter two are merely looking
for a single witness for having the property. Depending on the scenario,
specific properties of a collision may be of more interest than others.

The first metric evaluates the flexibility of the differential path. A
differential path with more flexibility will have more differentials which
produce blocks with the given differential path. More differentials implies
a greater flexibility in choice of colliding block, and possibly allowing
for multiple collisions for a given colliding block. Furthermore, a collision
with more differentials is more likely to satisfy the last metric, having a
pair of blocks--both ASCII--which produce a collision.

@@

The following techniques have been extensively tested on MD4 and partially
tested on MD5 and believe to apply fully to MD5. They likely apply to any
hash function with similar structure, but as of yet, are untested.

>>>>

The remainder of this paper is organized as follows. In Section
\ref{Sec:Related} we discuss previous results and show how our research
expands upon prior work. In Section \ref{Sec:Terminology}, we discuss the
terminology and notations we use throughout the remainder of the paper.
In Section \ref{Sec:Intuition}, we give the intuition behind measuring
the utility of a collision and motivate why it is useful. In Section
\ref{Sec:Logical}, we present new techniques in logical cryptanalysis to
analyze collisions in hash functions. In Section \ref{Sec:Empirical}, we
discuss the utility of existing collisions under the metrics described in
\ref{Sec:Intuition}. In Section \ref{Sec:Collision}, we discuss new
differential paths and the effectiveness of various techniques presented
in Section \ref{Sec:Logical}. Lastly, in Section \ref{Sec:Impacts}, we
discuss broader impacts of this research.

@@

Differential cryptanalysis has been the major technique behind the discovery
of collisions in cryptographic hash functions. Its importance can be seen
everywhere from X. Wang's attacks on MD4 \cite{Wang2005} to more recent
cryptanalysis of general purpose hash functions such as Murmur3 by
J. Aumasson, D. Bernstein, and M. Bo{\ss}let \cite{murmur3DC}. However, while
differential cryptanalysis plays an important role, techniques for automated
analysis of hash functions are still an emerging area. Further, techniques
are being developed to make hash functions impervious to differential
analysis.

@@

Logical cryptanalysis and its encoding as SAT likely started under the work of
F. Massacci in 1999 with his paper titled ``Using walk-SAT and rel-sat for
cryptographic key search''. \cite{Massacci:1999:UWR:1624218.1624261}. However,
much of the early work by F. Massacci was focused on symmetric and asymmetric
ciphers
\cite{Massacci:1999:UWR:1624218.1624261, MASSACCI2000LogicalCA, FIORINI2003101}.
It wasn't until the work of D. Jovanovi{\'{c}} and P. Jani{\v{c}}i{\'{c}}
that exploring collisions in the context of SAT was introduced
\cite{LogicalAnalysis}, and until the work of I. Mironov and L. Zhang that
this was studied for an existing collision class \cite{Mironov2006}.

@@

We make use of the \textit{bc2cnf} utility by T. Junttila as the basis for
our models \cite{circuits}. This allows us to encode hash functions as generic
structures in the circuit description language and ease the algorithmic
creation of models which build on top of them. We then use CryptoMiniSat5 by
M. Soos to run the models and check for a satisfying witness, often an
example collision \cite{Soos:2009:ESS:1575471.1575502, CryptoMiniSat5}. We
deviate from evaluating the performance of SAT solvers or encodings of models
by instead focusing on evaluating the properties of the hash function. We have
developed new techniques for using a SAT solver to inductively build full
collisions in MD4 from collisions in reduced-round MD4 and to build additional
examples of collision classes from a single instance of a collision in MD4.

@@

We can define a similar distance function, $\Delta$, between families of
collision classes by cardinality of the symmetric difference in the two
collision families. That is, given two families of collision classes,
$F_1, F_2 \in \mathcal{F}$:
\begin{align}
    \Delta(F_1, F_2) = \big| (F_1 \cup F_2) \setminus (F_1 \cap F_2) \big|
\end{align}
This is convenient for when the specifics of the differential path do not
matter, merely that there exists at least one collision class of the specified
form.

@@

Neighborhoods can be classified into three types: \textit{expansion},
\textit{internal}, and \textit{mixed}. Let $d$ be fixed. An \textit{expansion}
neighborhood of a collision class, $C \in \mathcal{C}$, is the neighborhood
restricted only to those collision classes which only differ in rounds external
to the collision family of $C$. An \textit{internal} neighborhood of $C$ is the
neighborhood restricted only to those collision classes which differ in rounds
internal to the collision family of $C$. A \textit{mixed} neighborhood of $C$
is the neighborhood restricted only to collisions which differ in a round
internal and a round external to the collision family of $C$. That is:
\begin{align*}
    N_{exp}(C, d) = \{ & C_{j} \in N(C, d) : \forall i \in F(C), C_j[i] = C[i] \} \\
    N_{int}(C, d) = \{ & C_{j} \in N(C, d) : \forall i \not\in F(C), C_j[i] = C[i] \} \\
    N_{mix}(C, d) = \{ & C_{j} \in N(C, d) : \exists i \in F(C), C_j[i] \neq C[i] \\
        & \text{ and } \exists k \not\in F(C), C_j[k] \neq C[k] \}
\end{align*}

@@

To begin, real world systems are most vulnerable to two attacks: the preimage
and second preimage. For systems relying on password validation via
comparing two hashes, preimage attacks would allow inverting the hash,
increasing the risk associated with losing a database of hashed passwords.
On the other hand, systems relying on message validation, such as signature
checks or file integrity checks would be easily broken with fast second
preimage techniques. Both of these attacks are intractable, thus protecting
real-world systems. However, much research has been devoted to collision
attacks, producing efficient results.

@@

Seeing the need to evaluate how much flexibility exists in a collision,
we propose the following metrics for evaluating the utility of a collision
class:
\begin{enumerate}
    \item The number of unique differentials a collision class has.
    \item The number of unit-step neighbors a collision class has.
    \item The maximum count of zeros in a the binary representation of a
        colliding block (and likewise with ones).
    \item Whether there exists a block which collides under multiple initial
        values.
    \item Whether or not zero, one, or both of the blocks in a collision may
        be of ASCII values under any input block difference.
\end{enumerate}
Note that the first three are quantitative measures providing some measure of
flexibility of a collision class, whereas the latter two are merely looking
for a single witness having such a property. Depending on the scenario,
specific properties of a collision may be of more interest than others.

The first metric evaluates the flexibility of the differential path. A
differential path with more flexibility will have more differentials which
produce blocks with the given differential path. More differentials implies
a greater flexibility in choice of colliding block, and possibly allows
for multiple collisions for a given colliding block. Furthermore, a collision
with more differentials is more likely to satisfy the last metric, having a
pair of blocks--both ASCII--which produce a collision.

@@

The following techniques have been extensively tested on MD4 and partially
tested on MD5 and are believed to apply fully to MD5. They likely apply to any
hash function with similar structure, but as of yet, are untested.
